Reformulating phylogenetic mixed models to improve flexibility and speed

### 2021-06-24 response.md

**We appreciate the opportunity to resubmit a revised version.**

As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).

**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.**

**BMB: I think if we have a non-problematic example we should include it**
**MLI: I think we cleaned up garamszegi's example for the package already. DOUBLE CHECK**

@@

A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it. For example, Kostikova et al. 2016 (DOI:10.1093/sysbio/syw010) present a model that seems relevant here. Further, claiming that ‘brms’ is more efficient than ‘MCMCglmm’ needs to be backed up by either a reference or a test (see also comments from the reviewers).

** MLi: This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation. **

In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.

**MLi: Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**

@@

- In the same spirit of applied statistics, a brief description (or table) of which functions to use in the libraries lme4 and glmmTMB to make these models work would make it easier to start. There should also be mention of the tutorial available in the repository

**MLi: we have now added a tutorial**

- It would also be beneficial to add an application of the method with biological data directly in the article. The more your article is associated with biological ideas, the more users you will reach in my opinion.

**Going to be a little annoying looking for an example with replicates.**
**Skip?**
**garamszegi's example is real**

### 2021-06-24 phyloglmm_ms_resubmit.tex
minor changes in the revision and response

\bmb{change PCM to PR (phylogenetic regression) throughout to narrow the scope? I made this change in the first para, would need to replace throughout if we think it's a good idea}
Phylogenetic regressions (PR) account for evolutionary relatedness when analyzing relationships among characteristics of species.
Given a known phylogeny, PRs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.

In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for PCM, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.

@@

In classic PCMs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}. 
If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.

Classic PCMs allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation leads to the  \emph{phylogenetic mixed model} \citep{housworth2004phylogenetic}.
For example, suppose we wish to fit a PCM to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
The standard PCM allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow to incorporate both
variation in brain size among taxa with similar body sizes (random intercepts) and
variation in the \emph{relationship} between predictors and responses among taxa (random slopes).

@@

We compare our approach with five other R packages that can fit PCMs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PCM models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\Sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.

@@

The range of PCMs presented here can incorporate many levels and types of phylogenetic variation.
Of course, there are further biological complexities we have neglected, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.
While it cannot easily incorporate these complexities, the approach here does offer an efficient way to analyze a wide range of evolutionary scenarios.
The general principle of encoding phylogenetic structure in the random effects model matrix can be implemented with any platform that supports continuous latent variables.
Our implementation allows users to explore new ideas by fitting phylogenetic mixed models with complex random effects to large data sets.

>>>>

Phylogenetic regressions (PR) account for evolutionary relatedness when analyzing relationships among characteristics of species.
Given a known phylogeny, PRs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.

In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.

@@

In classic PRs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}. 
If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.

Classic PRs allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PRs to allow for multiple sources of variation leads to the  \emph{phylogenetic mixed model} \citep{housworth2004phylogenetic}.
For example, suppose we wish to fit a PR to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
The standard PR allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow to incorporate both
variation in brain size among taxa with similar body sizes (random intercepts) and
variation in the \emph{relationship} between predictors and responses among taxa (random slopes).

@@

We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\Sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.

@@

The range of PRs presented here can incorporate many levels and types of phylogenetic variation.
Of course, there are further biological complexities we have neglected, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.
While it cannot easily incorporate these complexities, the approach here does offer an efficient way to analyze a wide range of evolutionary scenarios.
The general principle of encoding phylogenetic structure in the random effects model matrix can be implemented with any platform that supports continuous latent variables.
Our implementation allows users to explore new ideas by fitting phylogenetic mixed models with complex random effects to large data sets.
